Introduction to *cellcount* - Draft vignette

### 2023-05-01 cellcount_intro.Rmd
Re-formatting of document to 'distill' article, minor changes to documentation outline, and changes to text to reflect in-depth explanation of package materials

---
title: "Introduction to *cellcount* - Draft vignette"
author: "*T.E. Harman, M. Vandersea, K. Pokrzywinski*"
date: "`r Sys.Date()`"
abstract: >
  The ***cellcount*** package provides a simple approach to quantifying bacteria and cyanobacteria *via* flourescence microscopy. Microbial enumeration is an important tool for validating new methods and critical to assess potential ecological changes. However, standard manual microbial enumeration is time-consuming and notoriously difficult due to large colonies, filaments, and dense biomats. Conventional counting methods require specialized expertise and are error prone due to variations between researchers. Additionally, other methods of counting cells or particles (i.e. coulter counter, flow cytometer, etc.) are not suitable for filamentous or colonial morphotypes that can clog flow cells and apertures. Here, we outline a new open-source package, ***cellcount***, designed in the R computing language, to quickly and accurately count various microbial morphotypes. In this open access package, fluorescent microscope images are analyzed to generate particle counts and quantify overall cell density from samples at comparable concentrations as conventional counting methods.
output: html_document
---

@@

@@

***cellcount*** is an open-source R package distributed by NOAA NCCOS HAB-F on [GitHub](https://github.com/orgs/HABF-SDI-NCCOS-NOAA/repositories) and uses components in the RStudio IDE program. It is advised to use both the latest version of R and RStudio before beginning.

@@

- [***EBImage***](http://bioconductor.org/packages/devel/bioc/vignettes/EBImage/inst/doc/EBImage-introduction.html)
- [***tiff***](https://www.rforge.net/tiff/)
- [***pixmap***](https://cran.r-project.org/package=pixmap)
- [***raster***](https://cran.r-project.org/web/packages/raster/index.html)
- [***beepr***](https://cran.r-project.org/web/packages/beepr/index.html)

@@

```{r,include=FALSE}
greyscale <- function(x, contrast = 2) {
  x <- contrast * x
  x <- x[, , 1] + x[, , 2] + x[, , 3]
  x <- x / max(x)
  x <- normalize(x, inputRange = c(0.1, 0.75))
  return(x)
}
```
```{r}
grey_images <- lapply(img_transposed, greyscale, contrast = 1)
display(grey_images[[5]])
```

@@

Before images are analyzed in bulk, it is encouraged to visually 
analyze image results and determine if changes need to be made. This could include adjusting the threshold
from the "mapped" function, the contrast from the "greyscale" function 
(if used), or could involve making adjustments to the "single_cell_convert" or
"filamentous_convert" functions, such as the width and height of the rectangular 
window, the pixel area threshold (i.e., removes cells or other objects under a 
certain pixel area), or the tolerance and 'ext' variables. Observe initial 
results as indicated below and make adjustments where necessary:

```{r,include=FALSE}
border <- function(x, y) {
  c(x[1:y[1], 1], x[1:y[1], y[2]], x[1, 1:y[2]], x[y[1], 1:y[2]])
}
count_cells <- function(x) {
  countAll <- max(x)
  dims <- dim(x)
  border1 <- border(x, dims)
  ids <- unique(border1[which(border1 != 0)])
  countInner <- countAll - length(ids)
  return(countInner)
}
count_images <- function(image, normalize = TRUE, removeEdgeCells = TRUE) {
  if (removeEdgeCells) {
    dims <- dim(image)
    border1 <- border(image, dims)
    ids <- unique(border1[which(border1 != 0)])
    inner <- rmObjects(image, ids)
    EBImage::colorLabels(inner, normalize = normalize)
  } else {
    EBImage::colorLabels(image, normalize = normalize)
  }
}
single_cell_convert <- function(x, w = 17, h = 17, offset = 0.001, areathresh = 50, tolerance= 0.5, ext = 1) {
  image <- thresh(x, w = w, h = h, offset = offset)
  image1 <- fillHull(image)
  image2 <- watershed(distmap(image1), tolerance = tolerance, ext = ext)
  nf <- computeFeatures.shape(image2)
  nr <- which(nf[, "s.area"] < areathresh)
  image3 <- rmObjects(image2, nr)
  return(image3)
}
```
```{r}
imagesConverted <- single_cell_convert(imagesMapped[[5]], w = 10, h = 10, offset = 0.001, areathresh = 250, tolerance = 0.8, ext = 1)
final_img <- count_images(imagesConverted, normalize = T, removeEdgeCells = T)
display(final_img)
```

@@

Once necessary adjustments have been made, the following for-loop is used to
analyze images of interested in bulk, one at a time. In this example, using the provided demo *Microcystis* sp.
images, the for-loop uses the "single_cell_convert" function to produce unique 
color labels on cells within an image and to produce a final image showcasing 
these labels using the "count_images" function. Finally, "count_cells" is used 
to count the number of unique color labels (i.e., cells). Code has also been integrated in the for-loop
to add these count results to the 'Cell.Count' data frame made earlier, and 
to save each unique color label image to the "image_savdir" directory:

@@

In same cases, some images that were initially taken may be too poor of quality or contain
high background fluorescence, among other errors that may arise. The following code analyzes 
outliers from the produced data, within a range of two standard deviations 
above and below the median. The "outlier_detect" function tags which images are
out of this range and dictates a warning within our data frame, which is shown 
below. This alerts the end-user to perhaps re-image a particular sample, or
to investigate the error further:

@@

In order to calculate cell density of these images, the
"cell_density" function can be implemented, using equations from Wetzel and Likens, 1991. Generating the total amount of 
cells counted from these images is needed, where it can be added to the 'Cell.Count' data frame using the following code:

@@

The average number of cells per image can also be calculated, as well as cells
per mL using the following code. Note that 'cell.av' must be divided by the
number of images the end-user has analyzed, and the 'cell.mL' must be divided
by the total volume from the original sample:

@@

Informal discussions with researchers have expressed interest in a easy-to-use analysis interface for inexperienced R users
to utilize the ***cellcount*** tool. The following overview uses ***shiny*** apps to enter input data and run image analyses. Several packages are required to utilize these UI tools:

- [***shiny***](https://cran.r-project.org/web/packages/shiny/index.html)
- [***shinyalert***](https://cran.r-project.org/web/packages/shinyalert/index.html)
- [***shinyFiles***](https://cran.r-project.org/web/packages/shinyFiles/index.html)
- [***shinyjs***](https://cran.r-project.org/web/packages/shinyjs/index.html)
- [***DT***](https://cran.r-project.org/web/packages/DT/index.html)
- [***shinyBS***](https://cran.r-project.org/web/packages/shinyBS/index.html)

@@

@@

# Literature cited:

Wetzel, RG and Likens, GE. 1991. Limnological Analyses, Second Edition. Springer-
Verlag. 391 pp.

Pau, G, Fuchs, F, Sklyar, O, Boutros, M, Huber, W. 2010. EBImage—an R package for image processing with applications to cellular phenotypes, Bioinformatics, 26(7): 979–981, https://doi.org/10.1093/bioinformatics/btq046

Pokrzywinski, K, Boyd, B, Smith, J. 2019. A High-throughput Method for Counting 
Cyanobacteria Using Fluorescence Microscopy. ERDC TR-19-21.

>>>>

---
title: "Introduction to *cellcount* - Draft vignette"
author: 
  - name: Tyler E. Harman 
    affiliation: Consolidated Safety Services, Fairfax, VA 22031, USA
      National Oceanic and Atmospheric Administration, National Centers for Coastal Ocean Science, Beaufort, NC 28516, USA
    affiliation_email: tyler.harman@noaa.gov
    orcid_id: 0000-0001-5007-8170

  - name: Mark W. Vandersea
    affiliation: National Oceanic and Atmospheric Administration, National Centers for Coastal Ocean Science, Beaufort, NC 28516, USA

  - name: David E. Berthold
    affiliation: Department of Biological Sciences, Florida International University, Miami, FL 33199, USA

  - name: H. Dail Laughinghouse
    affiliation: Department of Biological Sciences, Florida International University, Miami, FL 33199, USA
    orcid_id: 0000-0003-1018-6948

  - name: Kaytee Pokrzywinski
    affiliation: National Oceanic and Atmospheric Administration, National Centers for Coastal Ocean Science, Beaufort, NC 28516, USA
    orcid_id: 0000-0003-1523-9930
date: "`r Sys.Date()`"
output: distill::distill_article
---

@@

# Summary:

The ***cellcount*** package provides a simple approach to quantifying bacteria and cyanobacteria *via* flourescence microscopy. Microbial enumeration is an important tool for validating new methods and critical to assess potential ecological changes. However, standard manual microbial enumeration is time-consuming and notoriously difficult due to large colonies, filaments, and dense biomats. Conventional counting methods require specialized expertise and are error prone due to variations between researchers. Additionally, other methods of counting cells or particles (i.e. coulter counter, flow cytometer, etc.) are not suitable for filamentous or colonial morphotypes that can clog flow cells and apertures. Here, we outline a new open-source package, ***cellcount***, a continuation of the report outlined in [@Pokrzywinski2019] designed in the R computing language, to quickly and accurately count various microbial morphotypes. In this open access package, fluorescent microscope images are analyzed to generate particle counts and quantify overall cell density from samples at comparable concentrations as conventional counting methods. In addition, additional tools are presented in this package using [R Shiny](https://cran.r-project.org/package=shiny) [@Chang2022] interfaces to demonstrate ease-of-use for researchers not familiar with the R environment. All materials outlined in this article, along with tutorials, are found on NOAA's NCCOS HAB-F [GitHub.](https://github.com/orgs/HABF-SDI-NCCOS-NOAA/repositories)

@@

***cellcount*** is an open-source R package distributed by NOAA NCCOS HAB-F on GitHub and uses components in the RStudio IDE program. It is advised to use both the latest version of R and RStudio before beginning.

@@

- ***EBImage*** [@Pau2010]
- ***tiff*** [@Urbanek2022]
- ***pixmap*** [@Bivand2021]
- ***raster*** [@Hijmans2023]
- ***beepr*** [@Baath2018]

@@

```{r,include=FALSE}
greyscale <- function(x, contrast = 2) {
  x <- contrast * x
  x <- x[, , 1] + x[, , 2] + x[, , 3]
  x <- x / max(x)
  x <- normalize(x, inputRange = c(0.1, 0.75))
  return(x)
}
```
```{r,fig.cap="\\label Converted greyscale image of *Microcystis* sp."}
grey_images <- lapply(img_transposed, greyscale, contrast = 1)
display(grey_images[[5]])
```

@@

Before images are analyzed in bulk, it is encouraged to visually 
analyze image results and determine if changes need to be made. This could include adjusting the threshold
from the *mapped* function, the contrast from the *greyscale* function 
(if used), or could involve making adjustments to the *single_cell_convert* or
*filamentous_convert* functions, such as the width and height of the rectangular 
window, the pixel area threshold (i.e., removes cells or other objects under a 
certain pixel area), or the tolerance and 'ext' variables. Observe initial 
results as indicated below and make adjustments where necessary:

```{r,include=FALSE}
border <- function(x, y) {
  c(x[1:y[1], 1], x[1:y[1], y[2]], x[1, 1:y[2]], x[y[1], 1:y[2]])
}
count_cells <- function(x) {
  countAll <- max(x)
  dims <- dim(x)
  border1 <- border(x, dims)
  ids <- unique(border1[which(border1 != 0)])
  countInner <- countAll - length(ids)
  return(countInner)
}
count_images <- function(image, normalize = TRUE, removeEdgeCells = TRUE) {
  if (removeEdgeCells) {
    dims <- dim(image)
    border1 <- border(image, dims)
    ids <- unique(border1[which(border1 != 0)])
    inner <- rmObjects(image, ids)
    EBImage::colorLabels(inner, normalize = normalize)
  } else {
    EBImage::colorLabels(image, normalize = normalize)
  }
}
single_cell_convert <- function(x, w = 17, h = 17, offset = 0.001, areathresh = 50, tolerance= 0.5, ext = 1) {
  image <- thresh(x, w = w, h = h, offset = offset)
  image1 <- fillHull(image)
  image2 <- watershed(distmap(image1), tolerance = tolerance, ext = ext)
  nf <- computeFeatures.shape(image2)
  nr <- which(nf[, "s.area"] < areathresh)
  image3 <- rmObjects(image2, nr)
  return(image3)
}
```
```{r,fig.cap="A converted color-label image of *Microcystis* sp."}
imagesConverted <- single_cell_convert(imagesMapped[[5]], w = 10, h = 10, offset = 0.001, areathresh = 250, tolerance = 0.8, ext = 1)
final_img <- count_images(imagesConverted, normalize = T, removeEdgeCells = T)
display(final_img)
```

@@

Once necessary adjustments have been made, the following for-loop is used to
analyze images of interested in bulk, one at a time. In this example, using the provided demo *Microcystis* sp.
images, the for-loop uses the *single_cell_convert* function to produce unique 
color labels on cells within an image and to produce a final image showcasing 
these labels using the *count_images* function. Finally, *count_cells* is used 
to count the number of unique color labels (i.e., cells). Code has also been integrated in the for-loop
to add these count results to the 'Cell.Count' data frame made earlier, and 
to save each unique color label image to the "image_savdir" directory:

@@

In same cases, some images that were initially taken may be too poor of quality or contain
high background fluorescence, among other errors that may arise. The following code analyzes 
outliers from the produced data, within a range of two standard deviations 
above and below the median. The *outlier_detect* function tags which images are
out of this range and dictates a warning within our data frame, which is shown 
below. This alerts the end-user to perhaps re-image a particular sample, or
to investigate the error further:

@@

In order to calculate cell density of these images, the
*cell_density* function can be implemented, using equations from [@Wetzel2000]. Generating the total amount of 
cells counted from these images is needed, where it can be added to the 'Cell.Count' data frame using the following code:

@@

The average number of cells per image can also be calculated, as well as cells
per mL using the following code. Note that 'cell.av' calculation must contain the
number of images the end-user has analyzed, and the 'cell.mL' must be divided
by the total volume from the original sample:

@@

Informal discussions with researchers have expressed interest in a easy-to-use analysis interface for inexperienced R users
to utilize the ***cellcount*** tool. The following overview uses 'Shiny' apps to enter input data and run image analyses. This results in the use of two functions, *input_data_GUI* and *analyze images_GUI*, to run the entire analysis process as outlined previously, and in the **sandbox_analysis_test.R** script. Several packages are required to utilize these UI tools:

- ***shiny*** [@Chang2022]
- ***shinyalert*** [@1Attali2021]
- ***shinyFiles*** [@Pedersen2022]
- ***shinyjs*** [@2Attali2021]
- ***DT*** [@Xie2023]
- ***shinyBS*** [@Bailey2022]

@@

Cells with a uniform chloroplast (auto-fluorescence) or fluorescent dyes illuminating the entire cell are the most useful when using this version of ***cellcount*** for enumeration purposes. Other cells such as dinoflagellates (i.e. *Karenia* sp.) that do not contain uniform chloroplasts may be more difficult to enumerate as the ***cellcount*** approach may separate these lobed chloroplasts into separate sections, causing enumeration outputs to be unreliable. In-situ hybridization dyes that target specific species or genera may be helpful if other cellular features are not useful or rather hinder in providing accurate enumeration.

@@

***cellcount*** provides an open-source approach to cyanobacteria enumeration that is accurate and rapid compared to standardized microscopy methods. The open-source approach allows researchers to access this tool for free and to modify and make additions the existing code to fulfill their particular needs, whereas the Shiny GUI allows users of all coding backgrounds to utilize this analysis tool.

# Literature Cited

---
references:
  - type: technical-report
    id: Pokrzywinski2019
    author:
      - family: Pokrzywinski
        given: K.
      - family: Boyd
        given: B.M.
      - family: Smith
        given: S.J.
    issued:
      date-parts:
      - - 2019
        - 8
    title: 'A high-throughput method for counting cyanobacteria using fluorescence microscopy'
    DOI: 10.21079/11681/33870
    URL: https://erdc-library.erdc.dren.mil/jspui/handle/11681/33870

  - type: article-journal
    id: Pau2010
    author:
      - family: Pau
        given: G.
      - family: Fuchs
        given: F.
      - family: Skylar
        given: O.
      - family: Boutros
        given: M.
      - family: Huber
        given: W.
    issued:
      date-parts:
      - - 2010
        - 3
        - 23
    title: 'EBImage—an R package for image processing with applications to cellular phenotypes'
    container-title: Bioinformatics
    volume: 26
    issue: 7
    page: 979-981
    DOI: 10.1093/bioinformatics/btq046
    URL: https://academic.oup.com/bioinformatics/article/26/7/979/211835

  - type: package
    id: Urbanek2022
    author:
      - family: Urbanek
        given: S.
      - family: Johnson
        given: K.
    issued:
      date-parts:
      - - 2022
        - 1
        - 31
    title: 'tiff: Read and Write TIFF Images. R package version 0.1-11.'
    URL: https://www.rforge.net/tiff/

  - type: package
    id: Bivand2021
    author:
      - family: Bivand
        given: R.
      - family: Leisch
        given: F.
      - family: Maechler
        given: Martin
    issued:
      date-parts:
      - - 2021
        - 1
        - 29
    title: 'Bitmap Images/Pixel Maps, R Package Version 0.4.12.'
    URL: https://cran.r-project.org/package=pixmap

  - type: book
    id: Wetzel2000
    author:
      - family: Wetzel
        given: R.G.
      - family: Likens
        given: G.E.
    issued:
      date-parts:
      - - 2000
        - 3
        - 3
    title: 'Limnological Analyses - 3rd Edition'
    page: 271-274
    DOI: 10.1007/978-1-4757-3250-4
    URL: https://link.springer.com/book/10.1007/978-1-4757-3250-4

  - type: package
    id: Baath2018
    author:
      - family: Baath
        given: R.
      - family: Dobbyn
        given: A.
    issued:
      date-parts:
      - - 2018
        - 6
        - 4
    title: 'beepr: Easily Play Notification Sounds on any Platform'
    URL: https://cran.r-project.org/web/packages/beepr/index.html

  - type: package
    id: Hijmans2023
    author:
      - family: Hijmans
        given: R.J
      - family: van Etten
        given: J.
      - family: Sumner
        given: M.
      - family: Cheng
        given: J.
      - family: Baston
        given: D.
      - family: Bevan
        given: A.
      - family: Bivand
        given: R.
      - family: Busetto
        given: L.
      - family: Canty
        given: M.
      - family: Fasoli
        given: B.
      - family: Forrest
        given: D.
      - family: Ghosh
        given: A.
      - family: Golicher
        given: D.
      - family: Gray
        given: J.
      - family: Greenburg
        given: J.A.
      - family: Hiemstra
        given: P.
      - family: Hingee
        given: K.
      - family: Ilich
        given: A.
      - family: Karney
        given: C.
      - family: Mattiuzzi
        given: M.
      - family: Mosher
        given: S.
      - family: Naimi
        given: B.
      - family: Nowosad
        given: J.
      - family: Pebesma
        given: E.
      - family: Lamigueiro
        given: O.P.
      - family: Racine
        given: E.B.
      - family: Rowlingson
        given: B.
      - family: Shortridge
        given: A.
      - family: Venables
        given: B.
      - family: Wueest
        given: R.
    issued:
      date-parts:
      - - 2023
        - 1
        - 16
    title: 'raster: Geographic Data Analysis and Modeling'
    URL: https://cran.r-project.org/web/packages/raster/index.html

  - type: package
    id: Bailey2022
    author:
      - family: Bailey
        given: E.
    issued:
      date-parts:
      - - 2022
        - 4
        - 17
    title: 'shinyBS: Twitter Bootstrap Components for Shiny'
    URL: https://cran.r-project.org/web/packages/shinyBS/index.html

  - type: package
    id: Chang2022
    author:
      - family: Chang
        given: W.
      - family: Cheng
        given: J.
      - family: Allaire
        given: J.
      - family: Boutros
        given: M.
      - family: Huber
        given: W.
    issued:
      date-parts:
      - - 2022
        - 12
        - 15
    title: 'Shiny: Web application framework for R'
    URL: https://cran.r-project.org/web/packages/shiny/index.html

  - type: package
    id: Pedersen2022
    author:
      - family: Pedersen
        given: T.L.
      - family: Nijs
        given: V.
      - family: Schaffner
        given: T.
      - family: Nantz
        given: E.
    issued:
      date-parts:
      - - 2022
        - 8
        - 19
    title: 'shinyFiles: A Server-Side File System Viewer for Shiny'
    URL: https://cran.r-project.org/web/packages/shinyFiles/index.html

  - type: package
    id: 1Attali2021
    author:
      - family: Attali
        given: D.
      - family: Edwards
        given: T.
      - family: Wang
        given: Z.
    issued:
      date-parts:
      - - 2021
        - 12
        - 20
    title: 'shinyalert: Easily Create Pretty Popup Messages (Modals) in Shiny'
    URL: https://cran.r-project.org/web/packages/shinyalert/index.html

  - type: package
    id: 2Attali2021
    author:
      - family: Attali
        given: D.
    issued:
      date-parts:
      - - 2021
        - 12
        - 23
    title: 'shinyjs: Easily Improve the User Experience of Your Shiny Apps in Seconds'
    URL: https://cran.r-project.org/web/packages/shinyjs/index.html

  - type: package
    id: Xie2023
    author:
      - family: Xie
        given: Y.
      - family: Cheng
        given: J.
      - family: Tan
        given: X.
      - family: Allaire
        given: J.J.
      - family: Girlich
        given: M.
      - family: Ellis
        given: G.F.
      - family: Rauh
        given: J.
      - family: Reavis
        given: B.
      - family: Gersen
        given: L.
      - family: Szopka
        given: B.
      - family: Pickering
        given: A.
      - family: Holmes
        given: W.
      - family: Marttila
        given: M.
      - family: Quintero
        given: A.
      - family: Laurent
        given: S.
    issued:
      date-parts:
      - - 2023
        - 1
        - 17
    title: 'DT: A Wrapper of the JavaScript Library DataTables'
    URL: https://cran.r-project.org/web/packages/DT/index.html
---
